Destruction of metrizability in generalized inverse limits

### 2015-08-19 articles/destroying-metrizability-in-generalized-inverse-limits/article.tex
add my revisions

\title{Joint work on long u.s.c. inverse limits}

@@

As various recent papers have shown (e.g., \cite{char roe}, \cite{varagona}, \cite{vernon}), continuum theorists have begun to broaden their study of generalized inverse limits to the case where the factor spaces are indexed by sets other than the positive integers. This more inclusive approach opens many new avenues for research, and has already produced some interesting results. For example, Patrick Vernon showed that an inverse limit indexed by the set of all integers with a single upper semi-continuous bonding function can be homeomorphic to a 2-cell \cite{vernon}, whereas V. Nall previously showed this is impossible for such an inverse limit indexed by the positive integers alone \cite{nall 2cell}.
\

In another recent paper \cite{varagona}, S. Varagona generalized many previously known theorems (from, e.g., \cite{ingram mahavier} and \cite{nall connected}) to the case of u.s.c. inverse limits indexed by arbitrary totally ordered sets. As he showed, in that context, inverse limits with a single idempotent upper semi-continuous bonding function become especially important. Relatively little is known about the kind of continua that may be produced by such inverse limits, however.
\

@@

\

@@

We begin with some useful notation. Let $X$ be a non-empty compact Hausdorff space. Then we denote by $2^X$ the set of all non-empty compact subsets of $X$. We say a subset $A$ of $X$ is a \emph{continuum} if $A$ is non-empty, compact and connected. $C(X)$ denotes the set of all elements of $2^X$ that are continua.
\

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and interval subsets of the real line, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$
\

@@

If $X$, $Y$, and $Z$ are non-empty compact Hausdorff spaces and $f : X \rightarrow 2^{Y}$ and $g: Y \rightarrow 2^{Z}$ are u.s.c., then the composition $g \circ f : X \rightarrow 2^{Z}$ is the u.s.c. function given by $(g \circ f)(x) = \{z \in Z \ | \ $There exists $y \in Y$ such that $y \in f(x)$ and $z \in g(y)\}$. In the special case where $f: X \rightarrow 2^{X}$, we denote $f \circ f$ by $f^2$. If $f^2 = f$, then we say $f$ is \emph{idempotent}.
\

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ always denotes the identity on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\{X_{\alpha}, f_{\alpha \beta}, \kappa\}$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \kappa\}$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.
\

Because it can be difficult to check if a large collection of u.s.c. functions is exact, a much more tractable case would be that of an inverse limit system with a single idempotent bonding function $f$. That is, suppose $X$ is a non-empty compact Hausdorff space and $f: X \rightarrow 2^{X}$ is an idempotent u.s.c. function. If $X_{\alpha} = X$ for each $\alpha \in \kappa$ and $f_{\alpha \beta} = f$ for each $\alpha < \beta \in \kappa$, then $\{X_{\alpha}, f_{\alpha \beta}, \kappa\} = \{X, f, \kappa\}$ is an \emph{inverse limit system with the single bonding function $f$}, and we denote the inverse limit of the system as $\varprojlim \{X, f, \kappa\}$. (Of course, since $f$ is idempotent, the collection $\textbf{f}$ of bonding functions is automatically exact.) We note that, although the only continuous idempotent surjective function from a compact Hausdorff space $X$ to itself is the identity map, there are many non-trivial examples of u.s.c. idempotent surjective functions on $X$.
\

@@

We now focus on the case of an inverse limit with a single idempotent, surjective, u.s.c. bonding function $f: [0,1] \rightarrow C([0,1])$. The main goal of this paper is to prove that if such an $f$ is not the identity map, then the inverse limit $\varprojlim \{[0,1], f, \kappa\}$ is metric exactly when the index set $\kappa$ is a countable ordinal.
\

@@

So, $f(1) = [a,b]$ where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1)=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.

\end{proof}

@@

@@

\begin{proof}
If $f = \iota$, then of course the inverse limit is homeomorphic to $[0,1]$. So, suppose $f \ne \iota$. If $\kappa$ is countable, then because the inverse limit is a subspace of $[0,1]^{\kappa}$, it is metric. However, if $\kappa$ is uncountable, then by Theorem \ref{main theorem}, the inverse limit contains a copy of $\kappa + 1$ and is therefore non-metric.
\end{proof}

@@

>>>>

\parskip=0.5em

\title{Destruction of metrizability in generalized inverse limits}

@@

As various recent papers have shown (e.g., \cite{char roe}, \cite{varagona}, \cite{vernon}), continuum theorists have begun to broaden their study of generalized inverse limits to the case where the factor spaces are indexed by sets other than the positive integers. This more inclusive approach opens many new avenues for research, and has already produced some interesting results. For example, Patrick Vernon showed that an inverse limit indexed by the set of all integers with a single upper semi-continuous bonding function can be homeomorphic to a 2-cell \cite{vernon}, whereas V. Nall previously showed this is impossible for such an inverse limit indexed by the positive integers alone \cite{nall 2cell}.

In another recent paper \cite{varagona}, S. Varagona generalized many previously known theorems (from, e.g., \cite{ingram mahavier} and \cite{nall connected}) to the case of u.s.c. inverse limits indexed by arbitrary totally ordered sets. As he showed, in that context, inverse limits with a single idempotent upper semi-continuous bonding function become especially important. Relatively little is known about the kind of continua that may be produced by such inverse limits, however.

@@

@@

We begin with some useful notation. Let $X$ be a non-empty compact Hausdorff space. Then we denote by $2^X$ the set of all non-empty compact subsets of $X$. We say a subset $A$ of $X$ is a \emph{continuum} if $A$ is non-empty, compact and connected. $C(X)$ denotes the set of all elements of $2^X$ that are continua.

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and intervals on a linearly ordered space, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$

@@

If $X$, $Y$, and $Z$ are non-empty compact Hausdorff spaces and $f : X \rightarrow 2^{Y}$ and $g: Y \rightarrow 2^{Z}$ are u.s.c., then the composition $g \circ f : X \rightarrow 2^{Z}$ is the u.s.c. function given by $(g \circ f)(x) = \{z \in Z \ | \ $There exists $y \in Y$ such that $y \in f(x)$ and $z \in g(y)\}$. In the special case where $f: X \rightarrow 2^{X}$, we denote $f \circ f$ by $f^2$. If $f^2 = f$, then we say $f$ is \emph{idempotent}.

Let $\kappa$ be an ordinal. (We follow the convention of set theory in which an ordinal is equal to the set of its predecessors. For more detailed background on ordinals, see, for example, \cite{Kunen}.) Suppose that for each $\alpha \in \kappa$, $X_{\alpha}$ is a non-empty compact Hausdorff space and whenever $\alpha \le \beta \in \kappa$, the set-valued function $f_{\alpha \beta} : X_{\beta} \rightarrow 2^{X_{\alpha}}$ is u.s.c. (With this notation, $f_{\alpha \alpha}$ always denotes the identity on $X_{\alpha}$.) Suppose further that, whenever $\alpha \le \beta \le \eta \in \kappa$, we have $f_{\alpha \beta} \circ f_{\beta \eta} = f_{\alpha \eta}$. Then the collection $\textbf{f} = \{ f_{\alpha \beta} \ | \ \alpha \le \beta \in \kappa\}$ of u.s.c. functions is said to be \emph{exact}. In this case, we say $\<X_{\alpha}, f_{\alpha \beta}, \kappa\>$ is an \emph{inverse limit system}; the \emph{inverse limit}, $\varprojlim \<X_{\alpha}, f_{\alpha \beta}, \kappa\>$, of this system is $\{ \textbf{x} \in \prod_{\alpha \in \kappa} X_{\alpha} \ | \ x_{\alpha} \in f_{\alpha \beta}(x_{\beta})$ for all $\alpha \le \beta \in \kappa\}$. (In keeping with tradition, a boldface $\textbf{x}$ always denotes an element of $\prod_{\alpha \in \kappa} X_{\alpha}$; the $\alpha$th coordinate of such an $\textbf{x}$ will be denoted $\textbf{x}(\alpha)$.)  We call the u.s.c. functions $f_{\alpha \beta}$ the \emph{bonding functions} of the inverse limit, and each $X_{\alpha}$ is called a \emph{factor space} of the inverse limit.

Because it can be difficult to check if a large collection of u.s.c. functions is exact, a much more tractable case would be that of an inverse limit system with a single idempotent bonding function $f$. That is, suppose $X$ is a non-empty compact Hausdorff space and $f: X \rightarrow 2^{X}$ is an idempotent u.s.c. function. If $X_{\alpha} = X$ for each $\alpha \in \kappa$ and $f_{\alpha \beta} = f$ for each $\alpha < \beta \in \kappa$, then $\<X_{\alpha}, f_{\alpha \beta}, \kappa\> = \<X, f, \kappa\>$ is an \emph{inverse limit system with the single bonding function $f$}, and we denote the inverse limit of the system as $\varprojlim \<X, f, \kappa\>$. (Of course, since $f$ is idempotent, the collection $\textbf{f}$ of bonding functions is automatically exact.) We note that, although the only continuous idempotent surjective function from a compact Hausdorff space $X$ to itself is the identity map, there are many non-trivial examples of u.s.c. idempotent surjective functions on $X$; see Figure \ref{idempotentUsc} for examples.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{idempotent_usc.pdf}
\end{center}
\caption{Examples of u.s.c. idempotent surjective functions on [0,1]}
\label{idempotentUsc}
\end{figure}

@@

We now focus on the case of an inverse limit with a single idempotent, surjective, u.s.c. bonding function $f: [0,1] \rightarrow C([0,1])$. The main goal of this paper is to prove that if such an $f$ is not the identity map, then the inverse limit $\varprojlim \{[0,1], f, \kappa\}$ is metric exactly when the index set $\kappa$ is a countable ordinal.

@@

So, $f(1) = [a,b]$ where $[a,b] \subseteq [0,x)$ or $[a,b] \subseteq (x,1)$. If $[a,b] \subseteq (x,1)$, then since $a > x$, $f(a) \subseteq [0,a)$ while $f(a)\subseteq f([a,b])=f(f(1))=f(1)=[a,b]$, contradiction. On the other hand, if $[a,b] \subseteq [0,x)$, then since $b<x$, $f(b) \subseteq (b,1]$ while $f(b) \subseteq f([a,b])=f(f(1)=f(1)=[a,b]$, another contradiction. Since each possible case results in a contradiction, the proof is complete.
\end{proof}

@@

Recall that a \textit{Corson compact} space is a compact space which embeds in a $\Sigma$-product of real lines (see \cite{alster}). All compact metric spaces are embeddable in $\mathbb R^\omega$ and therefore Corson compact.

@@

\begin{proof}
If $f = \iota$, then of course the inverse limit is homeomorphic to $[0,1]$. So, suppose $f \ne \iota$. If $\kappa$ is countable, then because the inverse limit is a subspace of $[0,1]^{\kappa}$, it is metric. However, if $\kappa$ is uncountable, then by Theorem \ref{main theorem}, the inverse limit contains a copy of the non-Corson compact space $\kappa + 1$.
\end{proof}

@@

\bibitem{alster} Alster, K., and Pol, Roman. "On function spaces of compact subspaces of $\Sigma$-products of the real line." Fundamenta Mathematicae 107.2 (1980): 135-143.
